---
title: "In-class Exercise 05: Visual Analytics for Building Better Explanatory Models"
author: "Ou Yiming"
date-modified: "last-modified"
format: 
  html:
    code-overflow: wrap
execute:
  echo: true
  eval: true
  warning: false
---

# Getting Started

Load the required packages.

::: callout-tip
## Tip for loading packages

`p_load` will first check if you have the package installed; otherwise, it will install it. Next, it will load the installed packages. Preferably, install the packages via `Tools > Install Packages` in RStudio before running `p_load`.
:::

This section loads necessary libraries for data manipulation, visualization, exploratory data analysis, and model performance evaluation.

```{r}
pacman::p_load(tidyverse, readxl, SmartEDA, easystats, gtsummary, ggstatsplot)
```

## Importing Data

This code imports the dataset from an Excel file.

```{r}
car_resale <- read_xls("data/ToyotaCorolla.xls", sheet = "data")
```

### Glimpse of the dataset

Displays the structure of the dataset including column names, data types, and values.

::: callout-note
This provides a quick overview of the dataset structure.
:::

::: panel-tabset
## Code

```{r}
#| eval: false
glimpse(car_resale)
```

## Output

```{r}
#| echo: false
glimpse(car_resale)
```
:::

### Listing the dataset

Lists all objects within the dataset.

::: callout-note
This helps to see the elements stored in the dataset.
:::

::: panel-tabset
## Code

```{r}
#| eval: false
list(car_resale)
```

## Output

```{r}
#| echo: false
list(car_resale)
```
:::

### Summary of the dataset

Generates summary statistics for each column in the dataset.

::: callout-note
Summary provides key statistical insights into each variable.
:::

::: panel-tabset
## Code

```{r}
#| eval: false
summary(car_resale)
```

## Output

```{r}
#| echo: false
summary(car_resale)
```
:::

### Overview of the data using SmartEDA

Displays an overview and structure of the dataset using SmartEDA.

::: callout-note
**Type 1** = Overview of Data \| **Type 2** = Structure of Data
:::

::: panel-tabset
## Code

```{r}
#| eval: false
car_resale %>% ExpData(type=1)
```

## Output

```{r}
#| echo: false
car_resale %>% ExpData(type=1)
```
:::

::: panel-tabset
## Code

```{r}
#| eval: false
car_resale %>% ExpData(type=2)
```

## Output

```{r}
#| echo: false
car_resale %>% ExpData(type=2)
```
:::

## Converting Columns to Factors

This section defines categorical columns and converts them into factors.

::: callout-tip
Converting to factors is essential for categorical analysis.
:::

```{r}
cols <- c("Mfg_Month", "HP_Bin", "CC_bin", "Doors", "Gears", 
          "Cylinders", "Fuel_Type", "Color", "Mfr_Guarantee", "BOVAG_Guarantee", 
          "Met_Color", "Automatic_airco", "Boardcomputer", "ABS", "Airbag_1", 
          "Airbag_2", "Airco", "CD_Player", "Central_Lock", "Powered_Windows", 
          "Power_Steering", "Radio", "Mistlamps", "Sport_Model", "Backseat_Divider", 
          "Metallic_Rim", "Radio_cassette", "Tow_Bar")

car_resale <- car_resale %>%
  mutate(Id = as.character(Id)) %>%
  mutate(across(all_of(cols), as.factor))
```

## Numerical Visualization

This section generates visualizations for numeric data with and without target variables.

::: callout-tip
## For more details

Please visit [SmartEDA Documentation](https://daya6489.github.io/SmartEDA/reference/ExpNumViz.html)
:::

::: panel-tabset
## Code

```{r}
#| eval: false
car_resale %>% ExpNumViz(target = NULL, nlim = 10, Page = c(2,2))
```

## Output

```{r}
#| echo: false
car_resale %>% ExpNumViz(target = NULL, nlim = 10, Page = c(2,2))
```
:::

::: panel-tabset
## Code

```{r}
#| eval: false
car_resale %>% ExpNumViz(target = "Price", nlim = 10, Page = c(2,2))
```

## Output

```{r}
#| echo: false
car_resale %>% ExpNumViz(target = "Price", nlim = 10, Page = c(2,2))
```
:::

::: panel-tabset
## Code

```{r}
#| eval: false
car_resale %>% ExpCatViz(target = NULL, col = "sky blue", clim = 10, margin = 2, Page = c(4,4), sample = 16)
```

## Output

```{r}
#| echo: false
car_resale %>% ExpCatViz(target = NULL, col = "sky blue", clim = 10, margin = 2, Page = c(4,4), sample = 16)
```
:::

## Multiple Linear Regression Method

Builds a multiple linear regression model with selected predictors.

::: callout-tip
Linear regression is used to understand relationships between the dependent variable and predictors.
:::

```{r}
model <- lm(Price ~ Age_08_04 + Mfg_Year + KM + Weight + Guarantee_Period,
            data = car_resale)
model
```

## Check for multicollinearity

Checks for multicollinearity between predictors and plots the Variance Inflation Factor (VIF).

::: callout-tip
For more details, please visit [easystats Performance Documentation](https://easystats.github.io/performance/reference/index.html)
:::

```{r}
check_collinearity(model)
```

```{r}
check_c <- check_collinearity(model)
plot(check_c)
```

## Checking for normality

Performs normality and heteroscedasticity checks on the regression model.

::: callout-tip
Normality and homoscedasticity are essential assumptions for linear regression.
:::

```{r}
model1 <- lm(Price ~ Age_08_04 + KM + Weight + Guarantee_Period, data = car_resale)
check_normality(model1)
```

```{r}
check_heteroscedasticity(model1)
```

::: panel-tabset
## Code

```{r}
check_model(model1)
```

## Output

```{r}
check_model(model1)
```
:::

## Model Summary and Regression Table

Displays model summary and regression table with fit statistics.

::: panel-tabset
## Code

```{r}
summary(model1)
```

## Output

```{r}
summary(model1)
```
:::

::: panel-tabset
## Option 1

```{r}
tbl_regression(model1, intercept = TRUE)
```

## Option 2

```{r}
tbl_regression(model1, intercept = TRUE) %>%
  add_glance_source_note(
    label = list(sigma ~ "σ"),
    include = c(r.squared, adj.r.squared, AIC, statistic, p.value, sigma)
  )
```
:::

```{r}
p_model1 <- parameters(model1)
```

::: panel-tabset
## Option 1: ggplot

```{r}
plot(parameters(model1))
```

## Option 2: ggstatsplot

```{r}
ggcoefstats(model1,
            output = "plot")
```
:::

# References

-   Prof Kam Tin Seong (2025) Lecture 5. Physical version.
