---
title: "In-class Exercise 04"
author: "Ou Yiming"
date-modified: "last-modified"
execute:
  echo: true
  eval: true
  warning: false
  freeze: true
---

# Installing the required R packages

First, write a code chunk to install and launch the following R packages

```{r}
pacman::p_load(ggdist, ggridges, ggthemes,
               colorspace, haven, SmartEDA, 
               tidyverse, tidymodels,
               ggscatterstats)
```

## Importing Data

Code chunk below uses read_csv() of readr to import Exam_data.csv into r environment as a tibble data.frame.

```{r}
exam_data <- read_csv("data/Exam_data.csv")
```

# Comparing different plots

## Plotting a Boxplot for each class

```{r}
ggplot(data = exam_data,
       aes(x = ENGLISH,
           y = CLASS)) +
  scale_x_continuous(
    name = "English grades",
    expand = c(0, 0)
  ) +
  geom_boxplot()
```

## Plotting a Ridgeline plot for each class

```{r}
ggplot(data = exam_data,
       aes(x = ENGLISH,
           y = CLASS)) +
  geom_density_ridges(
    scale = 3,
    rel_min_height = 0.01,
    bandwidth = 3.4,
    fill = lighten("#7097BB", .3),
    color = "white"
  ) +
  scale_x_continuous(
    name = "English grades",
    expand = c(0, 0)
  ) +
  scale_y_discrete(name = NULL, expand = expansion(add = c(0.2, 2.6))) +
  theme_ridges()
```

## Plotting a half eye graph

```{r}
ggplot(data = exam_data, 
       aes(x = RACE, 
           y = ENGLISH)) +
  stat_halfeye(adjust = 0.5,
               justification = -0.2,
               .width = 0,
               point_colour = NA) +
  geom_boxplot(width = .20,
               outlier.shape = NA) +
  stat_dots(side = "left", 
            justification = 1.2, 
            binwidth = .5,
            dotsize = 2)
```
# References

-   Prof Kam Tin Seong (2025) https://r4va.netlify.app/chap01. Online version
